Performance Evaluation of Big Data Processing Strategies for Neuroimaging

### 2018-12-14 paper.tex
Final read, added more comments, added a few sentences in conclusion

% Talk about containers somewhere?

@@

%\todo{Lazy evaluation should be mentioned here, or elsewhere in the paragraph.
%It is currently not described.}

@@

Spark will spill RDD data to disk whenever the complete dataset does not fit in
memory. Moreover, as Spark transformations generate new RDDs and numerous 
transformations may occur in a single application, Spark implements a Least-Recently Used (LRU) eviction
policy. If an evicted RDD needs to be reused, Spark will recompute it using the
lineage data collected. As demonstrated in~\cite{freeman2014mapping}, caching significantly improves processing times of iterative algorithms where 
RDDs are reused. It can be of even greater importance if the RDD is costly to 
recompute.
%mention results in thunder paper

@@

%% Main concepts
% pipeline description
% detailed description relevant to the experiment
% how shuffle works
% how data locality is implemented
% spills to disk when dataset (and derived data) is too large for memory
% Lazy evaluation
% persistence (refer to Thunder)
% multiple schedulers: we focus on Spark Standalone
% overlay cluster: not considered here as this would be scheduling

%% Technical details
% data serialization to Java (we use Python as most neuroinformatics do)
% limitation in size of RDD element, due to java implementation of binary data
% memory management: don't give too much memory to executors or GC
%  many places recommend limiting to 5 cores/executor, no one explains why
% more on tuning?

@@

% i use 'popular' for both...
Nipype is a popular Python neuroimaging processing engine. It aims at being a 
solution for easily creating reproducible neuroimaging workflows. Although Nipype
does not employ any Big Data processing strategies, it provides access to numerous
plugins to schedulers found in most clusters readily available to researchers, 
such as the SunGrid Engine (SGE), Portable Batch System (PBS), Slurm and HTCondor\todo{cite}. It also includes its own scheduler, MultiProc,
for parallel processing on single nodes. Nipype also includes many built-in 
interfaces to commonly used neuroimaging tools to be incorporated within the 
workflows. Nipype's ability to easily parallelize workflows in researcher-available 
cluster setups, capture detailed provenance information necessary for reproducibility,
and allow users to easily integrate existing neuroimaging tools, make it preferable
to existing Big Data solutions, which would necessitate modification to achieve this.

@@

% pipeline description: map nodes and their limitations
% general description relevant to the experiment
% two plugins
% provenance
% file system caching
% memory management (bad?)
% Nipype only has data locality becausae we use multiproc. Data locality is emulated by manual decisions on scheduling.

@@

Data storage location is critical to the performance of Big Data 
applications on HPC clusters.
Data may reside in the engine memory, on 
a file system whose content reside in virtual memory (for instance 
tmpfs), on disks local to the processing node, or on a shared 
file system. Table~\ref{table:features} summarizes the Big Data 
strategies that can be used depending on the data location. In 
addition, lazy evaluation is available in Spark regardless of data 
location. The remainder of this Section explains this Table and 
provides related performance models.
\begin{table}
\centering
\begin{tabular}{c|cc}
   \rowcolor{headcolor}
    Data Location                 & In-Memory     & Data Locality        \\
    \rowcolor{headcolor}
                                  & Computing     &                     \\
                                  \hline          
In-memory                         & \cellcolor{green!25} Yes           & \cellcolor{green!25}Yes                      \\
tmpfs                             & \cellcolor{green!25} Yes           & \cellcolor{green!25}Yes                  \\
Local Disk                        & \cellcolor{orange!25}Page Caching  & \cellcolor{green!25}Yes                  \\
Shared File System                & \cellcolor{orange!25}Page Caching  & \cellcolor{red!25}No                
% data locality is somehow present in shared fs due to page caching again.
\end{tabular}
\caption{Big Data strategies on a shared HPC cluster.}
\label{table:features}
\end{table}

@@

% measured disk bandwidth with random I/O, very different from sequential

Storing data on local disks obviously enables data locality as tasks 
executed on the node where their input files reside do not require to 
transfer input or output data. However, in absence of a more specific 
file system such as HDFS to handle 
file replication across computing nodes, data locality comes at the price
of stringent scheduling restrictions, as tasks can only be scheduled to the
single node that contains their input data.
% this brings data locality but not at the level of HDFS because replication isn't there and 
% there is only 1 single option to schedule a task while preserving data locality.

@@

% In~\cite{saini2012performance} the Lustre file system is benchmarked on NASA applications.

@@

% Dedicated Compute nodes with SLURM
% Lustre server

%~ \begin{table}
%~ \centering
%~ \begin{tabular}{c|c}
%~ $\delta/\gamma$ & $\Delta/\Gamma$\\
%~ \hline
%~ 164.5~MB/s         & 3.2~MB/s
%~ \end{tabular}
%~ \label{table:infrastructure}
%~ \caption{Bandwidth thresholds on the target cluster using 15 worker nodes and a
         %~ maximum of 9 CPUs on each node.}
%~ \end{table}

@@

% BigBrain
% Some BIDS dataset for fmriprep
We used BigBrain~\cite{amunts2013bigbrain}, a 75GB 40$\mu$m isotropic 
histological image of a 65-year-old male's brain. The BigBrain was selected due 
to being the only one of its kind, as there does not yet exist any other
higher-resolution image of a human brain. Moreover, there currently exists a 
lack of standardized tools for processing the BigBrain as a consequence 
of its size. To examine the effects processing the BigBrain has on Big Data 
strategies, we partitioned the full $3845\times3015\times3470$ voxel image into 30 ($5\times3\times2$)
chunks, 125 ($5\times5\times5$) chunks and 750 ($5\times15\times10$) chunks. Additionally, the full 
image was also split in half ($769\times603\times347$ voxels) and the half image was 
partitioned into 125 chunks.

@@

% Incrementation (avoid cache effects in binarization)
In order to investigate how the different strategies impact processing, we 
selected a simple incrementation pipeline (Algorithm~\ref{alg:incrementation}) 
that consisted exclusively of map 
stages. A series of map-only stages would enable us to evaluate the effects of
in-memory computing when data locality is preserved. Incrementation was 
selected 
over other applications, such as binarization, as it ensured that a new image
was created at each step (i.e. no caching effects within the executing 
application). Each partitioned chunk was incremented by 1, in parallel, by
the applications. As incrementing the images is a relatively quick 
process, we added a sleep delay to the incrementation stage in order to 
increase task duration as required. The incremented chunks would be either 
maintained in-memory (Spark only) or saved to either tmpfs, local disk or 
Lustre (Spark and Nipype). Should more than a single iteration be requested, 
the incremented chunks would be incremented again and saved to the same storage. 
This would repeat until the number of requested iterations had elapsed. In all 
conditions, the first input chunks and final output chunks would be read from / written to  Lustre. 
We chose to perform our initial 
reads and final writes on Lustre as a multi-tenant clusters' compute node's 
local storage, if 
available, is generally temporary and only available for the duration of the 
allocation.

@@

% Interactive gantt charts?

@@

\subsection{Effect of Data Locality}
%~ % It should be important, in particular when there is contention and when
%~ % tasks are small. 
We measure the effect of data locality by comparing the runs of Spark
on local disk (non-hatched green bars in Fig.~\ref{fig:results}) to the 
ones of Spark on Lustre (non-hatched blue bars). The speed-up provided by 
local execution peaks at 3.2, reached at 750 chunks in experiment 3. Overall, 
writing locally was usually preferrable to writing to Lustre, as a result of the
lower contention on local disk. Although it may be true, as stated in~\cite{Anantharayanan2011datalocality},
network bandwidths generally exceed that of disks, locality remains important 
as contention to a shared filesystem will be much higher than local disk.\todo{This needs to be finished or removed.}
The only time in our experiments where it was found that writing locally did not
have a huge impact over Lustre, was in experiments 1 and 4, at 1 iteration and 
when processing the MRI image, respectively. In both these scenarios, the Lustre 
writes did not impact performance as the data was able to be written to page cache
and flushed to Lustre asynchronously.

@@

\subsection{Scheduling Remarks}
% Scheduling: load balance as much as you can.
% (which isn't what nipype multiproc does)

@@

@@

With Nipype, MapNodes, which apply a given function to each element in a list, 
were found to be slower than the Node, which apply a function to a single element,
due to a blocking mechanism. For this reason, we selected to iterate through a 
series of Nodes in our code.

@@

It is common in neuroimaging applications for users to want access to all intermediate
data. Such a feature is currently only possible when writing to shared filesystem. 
It would also not be an option with Spark in-memory. To enable this burst buffers or 
heteorogeneous storage managers (e.g. Triple-H~\cite{tripleH}) could be used to ensure
fast processing and that all outputs (including intermediate outputs) will be 
sent asynchronously to the shared filesystem. \todo{double check triple-H}

% fault-tolerance: if you loose a node, you loose the data when it's written on disk or in memory.
%                  then the only solution is to recompute, which spark would do automatically.
%                  when nipype is run on lustre, it would avoid recomputing existing files, which Spark doesn't do.
%                   this niype feature is lost when not using lustre.

% Comment on the practical implications of using tmpfs or local disk: we might not get all the intermediary files.

% mention burst buffers or heterogeneous storage managers as potential solutions to write to local disk, 
% i.e., to benefit from data locality while getting your results on Lustre.

% We haven't benchmarked shuffles but we know it's important.

\subsection{Engine-Specific Comparisons}

@@

% Spark vs Nipype
% Do we see the effect of Java serialization?

@@

Big Data performance optimization strategies help improve performance of typical 
neuroimaging applications. Our experiments indicate that overall, in-memory computing 
enables greater speedups than what can be obtained by using page cache and tmpfs. 
While page cache and tmpfs do give memory-like performance, they are likely to fill
up faster than memory, leading to increased performance penalties when compared
to in-memory computing.

Data-locality was also found to play an important role in 
application performance. Local disk was found to perform between than the shared
filesystem, despite having lower bandwidth, due to increased contention on the 
shared filesystem. As local disk will generally have less contention than a 
shared filesystem. It is recommended to have the data stored locally. Although, 
storing data locally may come at a cost to fault tolerance.

@@

Even though Big Data strategies are beneficial to the processing of Big 
neuroimaging data, it is estimated that it would require running a functional MRI
dataset with 200 concurrent subjects for any noticeable impact using our Lustre 
bandwidth estimate.

To further improve neuroimaging data processing times, scheduling strategies for 
Big Data application in a multitenant environment need to be investigated. 
Additionally, workflow-aware cache eviction policies that maximizes page cache 
usage for the workflows should be investigated.

% Summary of the achievements.
% When is it useful to use a Big Data engine for neuroinformatics applications?

% Future work:
% - now that we know that Spark brings performance to neuro applications, make it usable
% - scheduling in a shared environment.
% - benchmark a large scale fmri processing to confirm our speculations
% - try in-memory distributed file systems such as Apache Ignite
% - workflow-aware cache eviction strategies instead of LRU/n
% - improve Spark to better use page caches
%    "```Disk and network I/O, of course, play a part in Spark performance 
%    as well, but neither Spark nor YARN currently do anything to actively 
%    manage them.```"
%    http://blog.cloudera.com/blog/2015/03/how-to-tune-your-apache-spark-jobs-part-2/

>>>>

@@

@@

Spark will spill RDD data to disk whenever the complete dataset does not fit in
memory. Moreover, as Spark transformations generate new RDDs and numerous 
transformations may occur in a single application, Spark implements a Least-Recently Used (LRU) eviction
policy. If an evicted RDD needs to be reused, Spark will recompute it using the
lineage data collected. As demonstrated in~\cite{freeman2014mapping}, caching significantly improves processing times of iterative algorithms where 
RDDs are reused. It can be of even greater importance if the RDD is costly to 
recompute.

@@

@@

% i use 'popular' for both... widely-used (for Spark)?
Nipype is a popular Python neuroimaging processing engine. It aims at being a 
solution for easily creating reproducible neuroimaging workflows. Although Nipype
does not employ any Big Data processing strategies, it provides access to numerous
plugins to schedulers found in most clusters readily available to researchers, 
such as the SunGrid Engine (SGE), Portable Batch System (PBS), Slurm and HTCondor\todo{cite}. It also includes its own scheduler, MultiProc,
for parallel processing on single nodes. Nipype also includes many built-in 
interfaces to commonly used neuroimaging tools to be incorporated within the 
workflows. Nipype's ability to easily parallelize workflows in researcher-available 
cluster setups, capture detailed provenance information necessary for reproducibility,
and allow users to easily integrate existing neuroimaging tools, make it preferable
to existing Big Data solutions, which would necessitate modification to achieve this.

@@

@@

Data storage location is critical to the performance of Big Data 
applications on HPC clusters.
Data may reside in the engine memory, on 
a file system whose content reside in virtual memory (for instance 
tmpfs), on disks local to the processing node, or on a shared 
file system. Table~\ref{table:features} summarizes the Big Data 
strategies that can be used depending on the data location. In 
addition, lazy evaluation is available in Spark regardless of data 
location. The remainder of this Section explains this Table and 
provides related performance models.
\begin{table}
\centering
\begin{tabular}{c|cc}
   \rowcolor{headcolor}
    Data Location                 & In-Memory     & Data Locality        \\
    \rowcolor{headcolor}
                                  & Computing     &                     \\
                                  \hline          
In-memory                         & \cellcolor{green!25} Yes           & \cellcolor{green!25}Yes                      \\
tmpfs                             & \cellcolor{green!25} Yes           & \cellcolor{green!25}Yes                  \\
Local Disk                        & \cellcolor{orange!25}Page Caching  & \cellcolor{green!25}Yes                  \\
Shared File System                & \cellcolor{orange!25}Page Caching  & \cellcolor{red!25}No                
\end{tabular}
\caption{Big Data strategies on a shared HPC cluster.}
\label{table:features}
\end{table}

@@

Storing data on local disks obviously enables data locality as tasks 
executed on the node where their input files reside do not require to 
transfer input or output data. However, in absence of a more specific 
file system such as HDFS to handle 
file replication across computing nodes, data locality comes at the price
of stringent scheduling restrictions, as tasks can only be scheduled to the
single node that contains their input data.

@@

% KEEP THIS FOR WHEN WE LOOK AT SIMULATION

@@

@@

We used BigBrain~\cite{amunts2013bigbrain}, a 75GB 40$\mu$m isotropic 
histological image of a 65-year-old male's brain. The BigBrain was selected due 
to being the only one of its kind, as there does not yet exist any other
higher-resolution image of a human brain. Moreover, there currently exists a 
lack of standardized tools for processing the BigBrain as a consequence 
of its size. To examine the effects processing the BigBrain has on Big Data 
strategies, we partitioned the full $3845\times3015\times3470$ voxel image into 30 ($5\times3\times2$)
chunks, 125 ($5\times5\times5$) chunks and 750 ($5\times15\times10$) chunks. Additionally, the full 
image was also split in half ($769\times603\times347$ voxels) and the half image was 
partitioned into 125 chunks.

@@

In order to investigate how the different strategies impact processing, we 
selected a simple incrementation pipeline (Algorithm~\ref{alg:incrementation}) 
that consisted exclusively of map 
stages. A series of map-only stages would enable us to evaluate the effects of
in-memory computing when data locality is preserved. Incrementation was 
selected 
over other applications, such as binarization, as it ensured that a new image
was created at each step (i.e. no caching effects within the executing 
application). Each partitioned chunk was incremented by 1, in parallel, by
the applications. As incrementing the images is a relatively quick 
process, we added a sleep delay to the incrementation stage in order to 
increase task duration as required. The incremented chunks would be either 
maintained in-memory (Spark only) or saved to either tmpfs, local disk or 
Lustre (Spark and Nipype). Should more than a single iteration be requested, 
the incremented chunks would be incremented again and saved to the same storage. 
This would repeat until the number of requested iterations had elapsed. In all 
conditions, the first input chunks and final output chunks would be read from / written to  Lustre. 
We chose to perform our initial 
reads and final writes on Lustre as a multi-tenant clusters' compute node's 
local storage, if 
available, is generally temporary and only available for the duration of the 
allocation.

@@

@@

\subsection{Effect of Data Locality}

We measure the effect of data locality by comparing the runs of Spark
on local disk (non-hatched green bars in Fig.~\ref{fig:results}) to the 
ones of Spark on Lustre (non-hatched blue bars). The speed-up provided by 
local execution peaks at 3.2, reached at 750 chunks in experiment 3. Overall, 
writing locally was usually preferrable to writing to Lustre, as a result of the
lower contention on local disk. Although it may be true, as stated in~\cite{Anantharayanan2011datalocality},
network bandwidths generally exceed that of disks, locality remains important 
as contention to a shared filesystem will be much higher than local disk.
The only time in our experiments where it was found that writing locally did not
have a huge impact over Lustre, was in experiments 1 and 4, at 1 iteration and 
when processing the MRI image, respectively. In both these scenarios, the Lustre 
writes did not impact performance as the data was able to be written to page cache
and flushed to Lustre asynchronously.

@@

\subsection{Scheduling Remarks}

@@

Spark currently doesn't include any active management of disk I/Os or 
page caches. We believe that it would be beneficial to extend it toward 
this direction, to increase the performance of operations where local 
storage has to be used, such as disk spills or shuffles. For instance, 
workflow-aware cache eviction policies that maximizes page cache usage 
for the workflows could be investigated.

@@

With Nipype, MapNodes, which apply a given function to each element in 
a list, were found to be slower than the Node, which apply a function 
to a single element, due to a blocking mechanism. For this reason, we 
selected to iterate through a series of Nodes in our code \todo{while 
MapNodes might have been easier to use?}.

\todo{If time permits, it would be useful to say that we checked that scheduling between 
both engines was similar, and show two Gantt charts to justify that.}

@@

It is common in neuroimaging applications for users to want access to all intermediate
data. Such a feature is currently only possible when writing to shared filesystem. 
It would also not be an option with Spark in-memory. To enable this, burst buffers or 
heteorogeneous storage managers (e.g. Triple-H~\cite{tripleH}) could be used to ensure
fast processing and that all outputs (including intermediate outputs) will be 
sent asynchronously to the shared filesystem. \todo{double check triple-H}

@@

@@

Big Data performance optimization strategies help improve performance of typical 
neuroimaging applications. Our experiments indicate that overall, in-memory computing 
enables greater speedups than what can be obtained by using page cache 
and tmpfs. While page cache and tmpfs do give memory-like performance, 
they are likely to fill up faster than memory, leading to increased 
performance penalties when compared to in-memory computing. We conclude 
that extending Big Data processing engines to better support 
neuroimaging applications, including developing their provenance, 
fault-tolerance, and reproducibility features, is a worthwhile project.

Data-locality was also found to play an important role in application 
performance. Local disk was found to perform between than the shared 
filesystem \todo{between than the shared filesystem: what does it 
mean?}, despite having lower bandwidth, due to increased contention on 
the shared filesystem. As local disk will generally have less 
contention than a shared filesystem. It is recommended to have the data 
stored locally. Although, storing data locally may come at a cost to 
fault tolerance.

@@

Even though Big Data strategies are beneficial to the processing of 
large images, it is estimated that it would require running a 
functional MRI dataset with 280 concurrent subjects for any noticeable 
impact using our Lustre bandwidth estimate. Benchmarking Spark and 
Nipype using such a large fMRI dataset would be a relevant follow-up 
experiment, to test this hypothesis. It would also be useful to evaluate
other types of applications, such as ones containing data shuffling steps.

Finally, we plan to extend this study by including task scheduling 
strategies in a multi-tenant environment. We expect to observe 
important differences between Spark and Nipype, due to Spark's use of 
overlay scheduling. The impact of other Big Data technologies, such as 
distributed in-memory file systems~\cite{ignite}, could also be 
investigated.
